Speeding up eQTL scans in the BxD population using GPU

### 2019-07-24 paper.tex
minor modification

\title{Real Time Genome Scan Using GPU}

\author[$\ast$]{Xiaoqi Hu}
\author[$\ast$]{Hyeonju Kim}
\author[$\ast$]{Gregory Farage}
\author[$\dagger$]{Pjotr Prins}
\author[$\dagger$]{Robert W. Williams}
\author[$\S$]{Karl W. Broman}
\author[$\ast$]{Saunak Sen}

@@

\runningtitle{G3 Journal Template on Overleaf} % For use in the footer

@@

% Use the \equalcontrib command to mark authors with equal
% contributions, using the relevant superscript numbers
\equalcontrib{1}
\equalcontrib{2}

\correspondingauthoraffiliation{3}{Corresponding author: Please insert the affiliation correspondence address and email for the corresponding author. The corresponding author should be marked with the relevant number in the author list, as shown in the example.}
\vspace{-34pt}% Only used for adjusting extra space in the left column of the first page

@@

@@

\subsection{Acceleration Techniques}
We used various techniques to accelerate eQTL scans. 
\subsubsection{Multithreaded CPU operations}
% Used for multi-threaded loops whenever possible, eg. for element-wise operations. 
 Our goal is to build a backend for GeneNetwork that allows researcher to interact with data in real time. 
 Such requirement sets up the standard that the genome scan must finish within seconds.
 To bring out the best performance of CPU, we use multithreaded operations whenever possible. 
 Julia, our choice of programming language, provide simple yet safe syntax for multithreading. 
 It is simply done by adding \code{Threads.@thread} macro to indicate to Julia that the following \code{for} loop is multhreaded region. 
 Using \code{Threads.nthreads()} function can show the number of threads in julia, and the default number of threads we use is 16 threads.

@@

% OpenBLAS for multi-threaded matrix operations. 
Since our algorithm largely depends on matrix operations, it is natual to find the fastest way to achieve best result, regardless of computing platform. 
There are various matrix libraries available for CPU, such as gslBLAS and OpenBLAS. 
They target different hardware or use various techniques to get highly optimal results. 
The matrix multiplication in OpenBLAS is default multithreaded, therefore does not require extra coding effort to make the CPU version of matrix multiplication run in parallel.
We chose OpenBLAS as CPU computing library.

Matrix multiplication and element-wise operation are algorithmically free of data and function dependency, it is amendable to GPU's parallel computing power.
Julia provides various packages for GPU including CUDA, OpenCL bindings. 
Our chosen hardware for GPU is from Nvidia, which requires their propriatory library CUDA. 
Besides the hardware requirement, CUDA is also mature and has been recongnized in the scientific computing community. 
\textit{cuBLAS} library is a fast GPU implementation of the BLAS from Nvidia. 
For matrix operations on GPU, we used cuBLAS library.

@@

\begin{figure}[!htb]
	\centering
	\caption{Variation of GPU vs CPU speedup with matrix shape
	}
	\label{GPUCPUShape}
	\includegraphics[scale = 0.36]{figs/speedup.png}
\end{figure}

@@

\subsubsection{Julia language}
%Programming simplicity of interpreted languages with speed approaching compiled languages; GPU programming and linear algebra straightforward
In our experiment of matrix multiplication, Julia's speed is comparable to C/C++. 
However, the low learning curve, clean syntax, as well as support to GPU programming libraries such as CUDAnative and CuArrays affords much lower programming efforts compared to C/C++.
Compared to writting GPU functions in C/C++, writing in Julia is cleaner and easier, because it requires much less boiler plate code.
Below is some code snippets to demonstrate GPU code, calling library or writing custom kernel in Julia.

@@

\subsection{Benefit of using GPU}
In parallel computing, Amdahl's law is used to find the theoretical maximum speedup by improving a particular part of a program.  
For example, if a program took ten minutes for a serial processor, and a function that takes nine our of those ten minutes can be parallelized, then, the theoretical speedup, no matter how many processors are used, can not be more than ten times, because the minimum execution time of this program is one minute. 
Therefore, profiling the entire genome scan process is the prerequisit in order to begin the optimization. 
Often, profiling would consider space and time complexity. 
Our primary concern is the time taken by each function, therefore only timing information is considered in our profiling. 
We used Julia's built-in Sampling Profiler %TODO reference here.)
to find our target function for GPU because it is less intrusive compared to other profiling methods.

The genome scan process includes the following steps:
\begin{itemize}
	\item Calculate standadized matrix for input matrices.
	\item Get correlation matrix
	\item Caclulate LOD score
\end{itemize}
Our profiling result shows that step two and three takes up 90\% of time, and both involve parallelizable matrix operations. 
So they are our candidates for GPU accleration.
To make fair comparison for CPU and GPU, the timing shown in Table \ref{tab:speedup-table} is the total execution time for genome scan. 
The \textit{No Find Max LOD} method shows the timing when running the scan with \textit{CPU only} and \textit{CPU\&GPU}.
As the table shows, the CPU\&GPU combination did not show any improvement over the serial version (CPU only) in terms of time. 
In contrast, it is slower, whether using single or double precision for the data type.
We used a GPU profiler \textit{nvprof} %TODO reference here
to investigate what is the bottleneck of GPU. 
Result is, 98\% of time is spent on data transfer from GPU to CPU (device to host).
As shown in Figure \ref{MatrixMult}, the input matrices Y' and G are small compared to the output matrix R.
For BXD spleen dataset, the Y' and G matrix is about 40MB %TODO: double check on this number
, while R matrix is about 4GB.
Data offloading is the main bottleneck for our GPU implementation.

@@

\subsection{Software Availability}
The source code for this package is publically available on Bitbucket: https://bitbucket.org/Xiaoqi-hu/lm-gpu.
This repository contains three folders: \textit{data}, \textit{src} and \textit{test}. 
Using Julia language and running \code{test.jl} in \textit{test} folder will run our method using either spleen data and hippocampus data in \textit{data} folder.

@@

\subsection{Platform}
Hardware:
 \begin{itemize}
	\item CPU: Intel Xeon Gold 6148; 40 cores @ 2.40GHz, 192GB 
	\item GPU: Tesla V100-PCIE; 5120 Cores @ 1380 MHz, 16GB
\end{itemize}
Software: 
\begin{itemize}
	\item Programming environment: Julia; CentOS 7
	\item Libraries: CUDA v10.1 and cuBLAS; OpenBLAS
	\item Profilers: Julia Profiler Package; nvprof
\end{itemize}

@@

\bibliography{example-bibliography}

>>>>

\title{Speeding up eQTL scans in the BxD population using GPU}

\author[$\ast$]{Xiaoqi Hu}
\author[$\ast$]{Hyeonju Kim}
\author[$\ast$]{Gregory Farage}
\author[$\dagger$]{Pjotr Prins}
\author[$\dagger$]{Robert W. Williams}
\author[$\S$]{Karl W. Broman}
\author[$\ast$,1]{Saunak Sen}

@@

\runningtitle{Speeding up eQTL scans in the BxD population using GPU} % For use in the footer

@@

% Use the \equalcontrib command to mark authors with equal
% contributions, using the relevant superscript numbers
%\equalcontrib{1}
%\equalcontrib{2}

\correspondingauthoraffiliation{1}{Corresponding author: Saunak Sen, Department of Preventive Medicine, University of Tennessee Health Science Center, Memphis, TN. sen@uthsc.edu}
\vspace{-34pt}% Only used for adjusting extra space in the left column of the first page

@@

Compare matrix eQTL \citep{shabalin2012matrix}, GridLMM \citep{runcie2018fast}, Bolt-LMM \citep{loh2015efficient}

@@

\subsection{Acceleration Techniques}
We used various techniques to accelerate eQTL scans. 
\subsubsection{Multithreaded CPU operations}
% Used for multi-threaded loops whenever possible, eg. for element-wise operations. 
 Our goal is to build a backend for GeneNetwork that allows researcher to interact with data in real time. 
 Such requirement sets up the standard that the genome scan must finish within seconds.
 To bring out the best performance of CPU, we use multithreaded operations whenever possible. 
 Julia \citep{bezanson2017julia}, our choice of programming language, provide simple yet safe syntax for multithreading. 
 It is simply done by adding \code{Threads.@thread} macro to indicate to Julia that the following \code{for} loop is multhreaded region. 
 Using \code{Threads.nthreads()} function can show the number of threads in julia, and the default number of threads we use is 16 threads.

@@

% OpenBLAS for multi-threaded matrix operations. 
Since our algorithm largely depends on matrix operations, it is natual to find the fastest way to achieve best result, regardless of computing platform. 
There are various matrix libraries available for CPU, such as gslBLAS and OpenBLAS \citep{wang2013augem}. 
They target different hardware or use various techniques to get highly optimal results. 
The matrix multiplication in OpenBLAS is default multithreaded, therefore does not require extra coding effort to make the CPU version of matrix multiplication run in parallel.
We chose OpenBLAS as CPU computing library.

Matrix multiplication and element-wise operation are algorithmically free of data and function dependency, it is amendable to GPU's parallel computing power.
Julia provides various packages for GPU including CUDA \citep{Nickolls:2008:SPP:1365490.1365500}, OpenCL bindings. 
Our chosen hardware for GPU is from Nvidia, which requires their propriatory library CUDA. 
Besides the hardware requirement, CUDA is also mature and has been recongnized in the scientific computing community. 
\textit{cuBLAS} \citep{cublas} library is a fast GPU implementation of the BLAS from Nvidia. 
For matrix operations on GPU, we used cuBLAS library.

@@

\begin{figure}[!htb]
	\centering
	\caption{Variation of GPU vs CPU speedup with matrix shape
	}
	\label{GPUCPUShape}
	\includegraphics[scale = 0.35]{figs/speedup.png}
\end{figure}

@@

\subsubsection{Julia language}
%Programming simplicity of interpreted languages with speed approaching compiled languages; GPU programming and linear algebra straightforward
In our experiment of matrix multiplication, Julia's speed is comparable to C/C++. 
However, the low learning curve, clean syntax, as well as support to GPU programming libraries such as CUDAnative \citep{CUDAnative.jl-2018} and CuArrays affords much lower programming efforts compared to C/C++.
Compared to writting GPU functions in C/C++, writing in Julia is cleaner and easier, because it requires much less boiler plate code.
Below is some code snippets to demonstrate GPU code, calling library or writing custom kernel in Julia.

@@

\subsection{Benefit of using GPU}
In parallel computing, Amdahl's law is used to find the theoretical maximum speedup by improving a particular part of a program.  
For example, if a program took ten minutes for a serial processor, and a function that takes nine our of those ten minutes can be parallelized, then, the theoretical speedup, no matter how many processors are used, can not be more than ten times, because the minimum execution time of this program is one minute. 
Therefore, profiling the entire genome scan process is the prerequisit in order to begin the optimization. 
Often, profiling would consider space and time complexity. 
Our primary concern is the time taken by each function, therefore only timing information is considered in our profiling. 
We used Julia's built-in Sampling Profiler to find our target function for GPU because it is less intrusive compared to other profiling methods.

The genome scan process includes the following steps:
\begin{itemize}
	\item Calculate standadized matrix for input matrices.
	\item Get correlation matrix
	\item Caclulate LOD score
\end{itemize}
Our profiling result shows that step two and three takes up 90\% of time, and both involve parallelizable matrix operations. 
So they are our candidates for GPU accleration.
To make fair comparison for CPU and GPU, the timing shown in Table \ref{tab:speedup-table} is the total execution time for genome scan. 
The \textit{No Find Max LOD} method shows the timing when running the scan with \textit{CPU only} and \textit{CPU\&GPU}.
As the table shows, the CPU\&GPU combination did not show any improvement over the serial version (CPU only) in terms of time. 
In contrast, it is slower, whether using single or double precision for the data type.
We used a GPU profiler \textit{nvprof} \cite{nvprof}
to investigate what is the bottleneck of GPU. 
Result is, 98\% of time is spent on data transfer from GPU to CPU (device to host).
As shown in Figure \ref{MatrixMult}, the input matrices Y' and G are small compared to the output matrix R.
For BXD spleen dataset, the Y' and G matrix is about 40MB %TODO: double check on this number
, while R matrix is about 4GB. 
Data offloading is the main bottleneck for our GPU implementation.

@@

\subsection{Software Availability}
The source code for this package is publically available on Bitbucket: \url{https://bitbucket.org/Xiaoqi-hu/lm-gpu}.
This repository contains three folders: \textit{data}, \textit{src} and \textit{test}. 
Using Julia language and running \code{test.jl} in \textit{test} folder will run our method using either spleen data and hippocampus data in \textit{data} folder.

@@

\subsection{Platform}
Hardware:
 \begin{itemize}
	\item CPU: Intel Xeon Gold 6148; 40 cores @ 2.40GHz, 192GB 
	\item GPU: Tesla V100-PCIE; 5120 Cores @ 1380 MHz, 16GB
\end{itemize}
Software: 
\begin{itemize}
	\item Programming environment: Julia; CentOS 7
	\item Libraries: CUDA v10.1 and cuBLAS; OpenBLAS
	\item Profilers: Julia Profiler; nvprof
\end{itemize}

@@

LM-GPU provides a fast backend for GeneNetwork that is powered by GPU. 
It enables researchers to do real-time interaction with the data provided in GeneNetwork. 
Such possibility can help researchers experiment with different parameters, and results in insightful findings. 
LM-GPU takes advantage of GPU's massive parallel computing power and Julia's elegant syntax for quick prototyping and implementing. 
We are able to run genome scan for hippocampus data with over one million phenotypes in three seconds.

%TODO: limitations. One degree of freedom

%Further development of LM-GPU is

\bibliography{reference}
